Structured Synchronous Reactive Programming for Game Development \\ \Large{Case Study: On Rewriting Pingus from C++ to \CEU}

### 2017-05-24 paper.tex
...

In this work, we advocate structured synchronous reactive programming as an
expressive and productive alternative for game logic development.
We present a case study of rewriting Pingus from C++ to \CEU with the
contributions as follows:
%
\begin{itemize}
\item Applying idiomatic code in \CEU as alternative solutions for six selected
      behaviors in the game logic.
\item Presenting an in-depth qualitative analysis of the proposed solutions in
      comparison to the original implementations in C++.
\item Identifying five recurrent control-flow patterns that likely apply to
      other games:
        \emph{Finite State Machines},
        \emph{Continuation Passing},
        \emph{Dispatching Hierarchies},
        \emph{Lifespan Hierarchies},
        \emph{Signaling Mechanisms}.
\end{itemize}
%
A control-flow pattern is a recurring technique to describe dependencies and
explicit orders between statements (or groups of statements) in a program.
For instance, consider how a key press stimulus propagates through the game
entities and also what happens with them if the stimulus causes the end of the
game.
%
\CEU supports primitives that help describing complex control-flow
relationships in the game logic more concisely.
%
The rewriting process consisted of identifying sets of callbacks implementing
control flow in the game and translating them to \CEU using appropriate
structured constructs.
%
As an example, a double mouse click is characterized by a first click, followed
by a maximum amount of time, followed by a second click.
This behavior depends on different events (clicks and timers) which have to
occur in a particular order.
In C++, the implementation involves callbacks crossing reactions to successive
events which manipulate state variables explicitly.
%
Although we documented in detail two use cases for each of the five patterns,
in this work we only present six cases due to space constraints.
%
This work focuses on a qualitative analysis for the programming techniques
that we applied during the rewriting process.
Not all techniques result in reduction in LoCs (especially considering the
verbose syntax of \CEU), but have other effects such as eliminating shared
variables and dependencies between classes.

The rest of the paper is organized as follows:
Section~\ref{sec.codebase} gives an overview of the Pingus codebases in C++ and
\CEU and describes our approach to identify and rewrite the control flow in the
game.
Section~\ref{sec.pats} discusses six case studies in detail which are
categorized in five control-flow patterns.
Section~\ref{sec.related} discusses related work.
Section~\ref{sec.conclusion} concludes the paper.

@@

\begin{enumerate}
\item \emph{Finite State Machines}:
    Event occurrences map to transitions between states that trigger
    appropriate actions comprising the behavior of a game entity.
\item \emph{Continuation Passing}:
    The completion of a long-lasting activity may carry a continuation, i.e.,
    some action to execute next in the game.
\item \emph{Dispatching Hierarchies}:
    Entities form a dispatching hierarchy in which a container that receives a
    stimulus automatically forwards it to its managed children.
\item \emph{Lifespan Hierarchies}:
    Entities form a lifespan hierarchy in which a terminating container entity
    automatically destroys its managed children.
\item \emph{Signaling Mechanisms}:
    Entities often need to communicate and affect each other explicitly through
    signaling mechanisms, especially if there is no hierarchy relationship
    between them.
\end{enumerate}

@@

@@

TODO: descrever o que sao actions, exibir imagem da tela com menu de actions e
seta apontando para pingu sendo clicado e com acao recem ativada

@@

\begin{figure}[t]
\centering
\includegraphics[width=\columnwidth]{bomber-state}
\caption{ State machine for the \emph{Bomber animation} sequence.
\label{fig.bomber}
}
\end{figure}

@@

The \emph{Bomber action} explodes the clicked pingu, throwing particles around
and also destroying the terrain under its radius.%
\footnote{Bomber action animation: \url{github.com/an000/p/blob/master/README.md#2} }
%
We can model the explosion animation with a sequential state machine
(Figure~\ref{fig.bomber}) with actions associated to specific frames as
follows%
\footnote{State machine animation: \url{github.com/an000/p/blob/master/README.md#3} }:
%
\begin{enumerate}
\item 0th frame:  plays a "Oh no!" sound.
\item 10th frame: plays a "Bomb!" sound.
\item 13th frame: throws particles, destroys the terrain, and shows an
                  explosion sprite.
\item Game tick:  hides the explosion sprite.
\item Last frame: kills the pingu.
\end{enumerate}

@@

TODO: what is this?

@@

Continuation passing typically controls the overall structure of the game,
such as screen transitions in menus and level progressions.
%
\CEU uses the direct style techniques in 5 cases involving screen transitions:
the main menu, the level menu, the level set menu, the world map loop, and
the gameplay loop.
%
It also uses the same technique for the loop that switches the pingu actions
during gameplay.

@@

@@

In C++, the class \code{Bomber} declares a \code{sprite} member to handle its
animation frames (Figure~\ref{fig.bomber}).
%
The \code{Sprite} class is part of the game engine and knows how to update and
render itself.
However, the \code{Bomber} still has to respond to \code{update} and
\code{draw} requests from the game and forward them to the sprite
(ln. 11--13 and 15--18).
%
To understand how the \code{update} callback flows from the original
environment stimulus from the game down to the sprite, we need to follow a long
chain of 7 method dispatches (Figure~\ref{fig.hier}):
%
\begin{enumerate}
\item \code{ScreenManager::display} in the main game loop calls \code{update}.
\item \code{ScreenManager::update} calls \code{last\_screen->update} for the
      active game screen (a \code{GameSession} instance, considering the
      \code{Bomber}).
\item \code{GameSession::update} calls \code{world->update}.
\item \code{World::update} calls \code{obj->update} for each object in the
      world.
\item \code{PinguHolder::update} calls \code{pingu->update} for each pingu
      alive.
\item \code{Pingu::update} calls \code{action->update} for the active pingu
      action.
\item \code{Bomber::update} calls \code{sprite.update}.
      \code{Sprite::update} finally updates the animation frame.
\end{enumerate}
%
Each dispatching step in the chain is necessary considering the game
architecture:
%
\begin{itemize}
\item With a single assignment to \code{last\_screen}, we can easily deactivate
      the current screen and redirect all dispatches to a new screen.
\item The \code{World} class manages and dispatches events to all game
      entities, such as all pingus and traps, with the common interface
      \code{WorldObj}.
\item Since it is common to iterate only over the pingus (vs. all world
      objects), the container \code{PinguHolder} manages all pingus.
\item Since a single pingu can change between actions during lifetime, the
      \code{action} member decouples them with another level of indirection.
\item Sprites are part of the game engine and are reusable everywhere (e.g., UI
      buttons, world objects, etc.), so it is also convenient to decouple them
      from actions.
\end{itemize}
%
The \code{draw} callback flows through the same dispatching hierarchy until
reaching the \code{Sprite} class.

@@

@@

\subsection{Signaling Mechanisms}
\label{sec.pats.signaling}

\subsubsection{Case Study: Global Keys and the Options Menu}

%\begin{figure}[t]
%\centering
%\includegraphics[width=\columnwidth]{options-anim-opt}
%\caption{The \emph{Mouse Grab} configuration option.
%\label{fig.options}
%}
%\end{figure}

\begin{figure}[t]
\centering
\includegraphics[width=\columnwidth]{events}
\caption{Mutual dependency between signals.
\label{fig.events}
}
\end{figure}

\begin{figure*}[t]
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
void GlbEvt::on_button_press (<...>) {
  <...>
  switch (event.keysym.sym) {
    case K_g:
      if (keys[K_LCTRL] || keys[K_RCTRL]) {
        cfgmgr.set_grab(
          !cfgmgr.get_grab());
      }
      break;
    <...>
  }
}

///

boost::signal<void(bool)> on_grab_change;

void CfgMgr::set_grab (bool v) {
  <...>
  if (v != get_grab()) {
    <...>   // the actual "grab" effect
    on_grab_change(v);
  }
}

///

boost::signal<void (bool)> on_change;

void ChkBox::set (bool on, bool sndsig) {
  <...>   // switches the check box state
  if (sndsig) {
    on_change(on);
  }
}

///

typedef std::vector<boost::connection> Conns;
Conns conns;

OptMenu::OptMenu() :
  conns(),
  grab_box(),
  <...>
{
  grab_box = new ChkBox(<...>);
  conns.push_back(
    cfgmgr.on_grab_change.connect( std::bind(
      &ChkBox::set, grab_box, <...>, false) );
    )
  );
  conns.push_back(
    grab_box->on_change.connect( std::bind(
      &CfgMgr::set_grab, &cfgmgr, <...>) );
    )
  );
  <...>

}

OptMenu::~OptMenu() {
  for (Conns::iterator i=conns.begin();
             i!=conns.end();
             ++i) {
    (*i).disconnect();
  }
}

.
\end{lstlisting}
\centering\small{\ax Implementation in C++}
\end{minipage}
%
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
spawn do
  var _SDL_KeyboardEvent&& e;
  every e in SDL_KEYDOWN do
    var _u8&& keys = _SDL_GetKeyState(null);
    <...>
    if e:keysym.sym == K_g then
      if keys[K_LCTRL] or keys[K_RCTRL] then
        emit cfgmgr.go_grab(
            not {cfgmgr.get_grab()});
      end
    end
    <...>
  end
end

///

data CfgMgr with
  event bool go_grab;
end
var CfgMgr cfgmgr = val CfgMgr(_);

spawn do
  var bool v;
  every v in cfgmgr.go_grab do
    <...>   // the actual "grab" effect
  end
end

///

data IChkBox with
  var   bool is_on;
  event bool go_click;
end

code/await ChkBox (<...>) -> (var IChkBox box) -> FOREVER do
  box = val IChkBox(<...>);
  <...>
  par do
    every c.component.on_click do
      emit box.go_click(not box.is_on);
    end
  with
    loop do
      <...>   // switches the check box state
      box.is_on = await box.go_click;
    end
  end
end

///

code/await OptMenu <...> do
  <...>
  var& ChkBox b2 = <...>;
  spawn do
    par do
      var bool v;
      every v in cfgmgr.go_grab do
        emit b2.box.go_click(v);
      end
    with
      var bool v;
      every v in b2.box.go_click do
        emit cfgmgr.go_grab(v);
      end
    end
  end
  <...>
end
\end{lstlisting}
\centering\small{\bx Implementation in \CEU}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ TODO.
\label{lst.grab}
}
\end{figure*}

The \emph{mouse grab option} restricts the mouse movement to the game window
boundaries%
\footnote{Mouse grab animation: \url{github.com/an000/p/blob/master/README.md#6} }.
The option can be set anywhere in the game by pressing \emph{Ctrl-G}.
In addition, the \emph{Options menu} has a check box to toggle the
\emph{mouse grab option} with mouse clicks while still responding to
\emph{Ctrl-G} presses.
%
Figure~\ref{lst.grab} compares the implementations in C++ and \CEU.

The implementations in C++ and \CEU use a signalling mechanism to connect the
key presses, the check box, and a configuration manager that applies the
appropriate side effects in the game (i.e., restrict the mouse movement).
Figure~\ref{fig.events} illustrates how the mutual notifications create a 
dependency cycle between the configuration manager and the check box.

In C++, the class `GlbEvt` detects \emph{Ctrl-G} presses and invokes the
callback \code{config\_manager.set\_grab} (ln. 5--8).
%
The class `CfgMgr` uses a \code{boost::signal} (ln. 16) to notify the
application when the new configuration is applied (ln. 22).
%
The \code{if} enclosing the signal emission (ln. 20--23) breaks the dependency 
cycle of Figure~\ref{fig.events} and prevents an infinite execution loop.
%
The class \code{ChkBox} also uses a \code{boost::signal} (ln. 28) to notify
the application on changes (ln. 33).
%
Again, the \code{if} enclosing the signal emission (ln. 32--34) breaks the 
dependency cycle of Figure~\ref{fig.events} to avoid infinite execution.
%
The class \code{OptMenu} creates the dependency loop by connecting the two
signals.
%
The constructor binds
the signal \code{config\_manager.on\_grab\_change} to the callback
    method \code{grab\_box->set} (ln. 48--52),
and also
the signal \code{grab\_box->on\_change} to the callback method
    \code{config\_manager.set\_grab} (ln. 53--57).
This way, every time the \code{CfgMgr} signals
\code{on\_grab\_change} (ln. 22), the method \code{set} is implicitly called.
The same happens between the signal \code{on\_change} in the \code{ChkBox}
and the method \code{set\_grab} in the \code{CfgMgr} (ln. 18).
%
Note that the signal binding to call \code{ChkBox::set} (ln. 50) receives a
fixed value \code{false} as the last argument to prevent infinite execution
(ln. 30).
%
The \code{OptMenu} destructor (ln. 62--66) breaks the connections explicitly
when the \emph{Option screen} terminates.

In \CEU, a \emph{Ctrl-G} key press broadcasts the internal event
\code{config\_manager.go\_grab} to the application (ln. 8).
%
The configuration manager just needs to react to \code{go\_grab}
continuously to perform the \emph{grab} effect (ln. 25--27).
%
The \code{ChkBox} exposes the event \code{go\_click} for notifications in
both directions, i.e., from the abstraction to the application and
\emph{vice versa}:
%
The abstraction reacts to external clicks continuously (ln. 41--43) to
broadcast the event \code{go\_click} (ln. 47).
It also reacts continuously to \code{go\_click} in another line of execution
(ln. 45--48), which awakes from notifications from the first line of execution
or from the application.
%
The \code{OptMenu} connects the two events as follows:
%
The two loops in parallel handle the connections in opposite directions:
from the configuration manager to the check box (ln. 60--62);
and
from the check box to the configuration manager (ln. 65--67).
%
When the \emph{Option screen} terminates, the connections break automatically
since the body with the two loops is automatically aborted.
%
Note that the implementation in \CEU does not check event emits to break the
dependency cycle and prevent infinite execution.
Due to the stack-based execution for internal events in
\CEU~\cite{ceu.sensys13}, programs with mutually-dependent events do not
create infinite execution loops.

First-class internal events of \CEU provide some advantages in comparison to 
Boost signals:

\begin{itemize}
\item They use the same convenient syntax of external events (e.g.,
      \code{emit}, \code{await}, \code{every}, etc.).
\item They never create infinite dependency loops.
\item They do not require explicit unbinding.
\end{itemize}

The implementation in \CEU uses 39 events for internal communication defined in
23 files with over 200 invocations of \code{emit} and \code{await} spread in
over 50 files.

Internal events are used for resizing entities, broadcasting update and draw
requests, pausing parts of the game, triggering new pingu actions, signalling
collisions, signalling UI interactions, among many others cases.

@@

We promote the \emph{structured synchronous reactive} programming model of
\CEU for the development of games.
We present in-depth use cases categorized in 5 control-flow patterns applied to
\emph{Pingus} (an open-source \emph{Lemmings} clone) that likely apply to other
games.

>>>>

In this work, we advocate structured synchronous reactive programming as an
expressive and productive alternative for game logic development.
We present a case study of rewriting Pingus from C++ to \CEU with the
contributions as follows:
%
\begin{itemize}
\item Applying idiomatic code in \CEU as alternative solutions for six selected
      behaviors in the game logic.
\item Presenting an in-depth qualitative analysis of the proposed solutions in
      comparison to the original implementations in C++.
\item Identifying four recurrent control-flow patterns that likely apply to
      other games:
        \emph{Finite State Machines},
        \emph{Continuation Passing},
        \emph{Dispatching Hierarchies}, and
        \emph{Lifespan Hierarchies}.
        %\emph{Signaling Mechanisms}.
\end{itemize}
%
A control-flow pattern is a recurring technique to describe dependencies and
explicit orders between statements (or groups of statements) in a program.
For instance, consider how a key press stimulus propagates through the game
entities and also what happens with them if the stimulus causes the end of the
game.
%
\CEU supports primitives that help describing complex control-flow
relationships in the game logic more concisely.
%
The rewriting process consisted of identifying sets of callbacks implementing
control flow in the game and translating them to \CEU using appropriate
structured constructs.
%
As an example, a double mouse click is characterized by a first click, followed
by a maximum amount of time, followed by a second click.
This behavior depends on different events (clicks and timers) which have to
occur in a particular order.
In C++, the implementation involves callbacks crossing reactions to successive
events which manipulate state variables explicitly.
%
We only present six cases distributed in the four patterns due to space
constraints, but we actually documented in detail a total eleven cases in six
patterns (we omit the sixth pattern \emph{Signaling Mechanisms} entirely).
%
This work focuses on a qualitative analysis for the programming techniques
that we applied during the rewriting process.
Not all techniques result in reduction in LoCs (especially considering the
verbose syntax of \CEU), but have other effects such as eliminating shared
variables and dependencies between classes.

The rest of the paper is organized as follows:
Section~\ref{sec.codebase} gives an overview of the Pingus codebases in C++ and
\CEU and describes our approach to identify and rewrite the control flow in the
game.
Section~\ref{sec.pats} discusses six case studies in detail which are
categorized in four control-flow patterns.
Section~\ref{sec.related} discusses related work.
Section~\ref{sec.conclusion} concludes the paper.

@@

\begin{enumerate}
\item \emph{Finite State Machines}:
    Event occurrences map to transitions between states that trigger
    appropriate actions comprising the behavior of a game entity.
\item \emph{Continuation Passing}:
    The completion of a long-lasting activity may carry a continuation, i.e.,
    some action to execute next in the game.
\item \emph{Dispatching Hierarchies}:
    Entities form a dispatching hierarchy in which a container that receives a
    stimulus automatically forwards it to its managed children.
\item \emph{Lifespan Hierarchies}:
    Entities form a lifespan hierarchy in which a terminating container entity
    automatically destroys its managed children.
%\item \emph{Signaling Mechanisms}:
    %Entities often need to communicate and affect each other explicitly through
    %signaling mechanisms, especially if there is no hierarchy relationship
    %between them.
\end{enumerate}

@@

    Event occurrences map to transitions between states that trigger
    appropriate actions comprising the behavior of a game entity.

@@

@@

\begin{figure}[t]
\centering
\includegraphics[width=120px]{bomber-01}
\caption{ Assigning the \emph{Bomber action} to a pingu.
\label{fig.bomber.action}
}
\end{figure}

\begin{figure}[t]
\centering
\includegraphics[width=\columnwidth]{bomber-state}
\caption{ State machine for the \emph{Bomber animation} sequence.
\label{fig.bomber.fsm}
}
\end{figure}

@@

In Pingus, the player may assign actions to specific pingus, as illustrated in
Figure~\ref{fig.bomber}.
%
The \emph{Bomber action} explodes the clicked pingu, throwing particles around
and also destroying the terrain under its radius.%
\footnote{Bomber action animation: \url{github.com/an000/p/blob/master/README.md#2} }
%
We can model the explosion animation with a sequential state machine
(Figure~\ref{fig.bomber.fsm}) with actions associated to specific frames as
follows%
\footnote{State machine animation: \url{github.com/an000/p/blob/master/README.md#3} }:
%
\begin{enumerate}
\item 0th frame:  plays a "Oh no!" sound.
\item 10th frame: plays a "Bomb!" sound.
\item 13th frame: throws particles, destroys the terrain, and shows an
                  explosion sprite.
\item Game tick:  hides the explosion sprite.
\item Last frame: kills the pingu.
\end{enumerate}

@@

    The completion of a long-lasting activity may carry a continuation, i.e.,
    some action to execute next in the game.

@@

Continuation passing typically controls the overall structure of the game,
such as screen transitions in menus and level progressions.
%
\CEU uses the direct style techniques in five cases involving screen
transitions:
the main menu, the level menu, the level set menu, the world map loop, and
the gameplay loop.
%
It also uses the same technique for the loop that switches the pingu actions
during gameplay.

@@

    Entities form a dispatching hierarchy in which a container that receives a
    stimulus automatically forwards it to its managed children.

@@

In C++, the class \code{Bomber} declares a \code{sprite} member to handle its
animation frames (Figure~\ref{fig.bomber.fsm}).
%
The \code{Sprite} class is part of the game engine and knows how to update and
render itself.
However, the \code{Bomber} still has to respond to \code{update} and
\code{draw} requests from the game and forward them to the sprite
(ln. 11--13 and 15--18).
%
To understand how the \code{update} callback flows from the original
environment stimulus from the game down to the sprite, we need to follow a long
chain of 7 method dispatches (Figure~\ref{fig.hier}):
%
\begin{enumerate}
\item \code{ScreenManager::display} in the main game loop calls \code{update}.
\item \code{ScreenManager::update} calls \code{last\_screen->update} for the
      active game screen (a \code{GameSession} instance, considering the
      \code{Bomber}).
\item \code{GameSession::update} calls \code{world->update}.
\item \code{World::update} calls \code{obj->update} for each object in the
      world.
\item \code{PinguHolder::update} calls \code{pingu->update} for each pingu
      alive.
\item \code{Pingu::update} calls \code{action->update} for the active pingu
      action.
\item \code{Bomber::update} calls \code{sprite.update}.
      \code{Sprite::update} finally updates the animation frame.
\end{enumerate}
%
Each dispatching step in the chain is necessary considering the game
architecture:
%
\begin{itemize}
\item With a single assignment to \code{last\_screen}, we can easily deactivate
      the current screen and redirect all dispatches to a new screen.
\item The \code{World} class manages and dispatches events to all game
      entities, such as all pingus and traps, with the common interface
      \code{WorldObj}.
\item Since it is common to iterate only over the pingus (vs. all world
      objects), the container \code{PinguHolder} manages all pingus.
\item Since a single pingu can change between actions during lifetime, the
      \code{action} member decouples them with another level of indirection.
\item Sprites are part of the game engine and are reusable everywhere (e.g., UI
      buttons, world objects, etc.), so it is also convenient to decouple them
      from actions.
\end{itemize}
%
The \code{draw} callback flows through the same dispatching hierarchy until
reaching the \code{Sprite} class.

@@

    Entities form a lifespan hierarchy in which a terminating container entity
    automatically destroys its managed children.

@@

@@

We promote the \emph{structured synchronous reactive} programming model of
\CEU for the development of games.
We present in-depth use cases categorized in four control-flow patterns applied
to \emph{Pingus} (an open-source \emph{Lemmings} clone) that likely apply to
other games.
